### 2022-03-30 manuscript.tex
today [ci skip]
This build is based on
https://github.com/scherzerthesis/thesis/commit/cacf67ddbb3ecfa8a81d823abdfbeafd930f1848.

This commit was created by the following CI build and job:
https://github.com/scherzerthesis/thesis/commit/cacf67ddbb3ecfa8a81d823abdfbeafd930f1848/checks
https://github.com/scherzerthesis/thesis/actions/runs/2066363553

This manuscript
(\href{https://scherzerthesis.github.io/thesis/v/330b3032e45afe681efd66fc81301108e295a421/}{permalink})
was automatically generated
from \href{https://github.com/scherzerthesis/thesis/tree/330b3032e45afe681efd66fc81301108e295a421}{scherzerthesis/thesis@330b303}
on March 29, 2022.

@@

\emph{\textbf{P53-mutant Alleles Differentially Alter BRAF (V600E) Lung Tumorigenesis}}

@@

\emph{\textbf{Mutant-p53 Accumulation is Associated with Dedifferentiation and p19 Induction}}

Increased malignancy in lung adenocarcinoma is often associated with altered states of tumor cell identity and dedifferentiation (van Veen \emph{et al.}, 2019; Marjanovic \emph{et al.}, 2020). Since LUADs often arise from AT2 cells, we surveyed the expression of AT2 cell identifiers, such as NKX2-1, a master regulator of AT2 transcriptional programs, and SPC, a marker of differentiated AT2 cells (Marjanovic \emph{et al.}, 2020). Quantification of over 50 tumors from \emph{Braf\textsuperscript{V600E/+}} mice with either \emph{Trp53\^{} flfl\^{}}, \emph{Trp53\textsuperscript{R172H/R172H}}, and \emph{Trp53\textsuperscript{R245W/R245W}} revealed that \emph{Trp53\textsuperscript{R172H/R1722H}} mice had significantly more tumors that stained negative for NKX2-1 and SPC (Figure \ref{fig:12} A an B). This finding strongly indicates that mutant-p53 tumors are dedifferentiated. Since HMGA2 expression is associated with LUAD dedifferentiation and worse patient prognosis, we performed IHC for HMGA2 (Cite). We noted an increase in HMGA2-positive tumors in both \emph{Trp53\textsuperscript{R172H/R172H}}, and \emph{Trp53\textsuperscript{R245W/R245W}} tumors compared to \emph{Trp53\textsuperscript{fl/fl}} (Figure \ref{fig:12} A,B). Thus, this data taken together reveals that Trp53R172H/R172H mice exhibit both a higher tumor burden and worse overall survival than \emph{Trp53\textsuperscript{fl/fl}}, but the tumors developed in these mice are more likely to exhibit malignant progression and dedifferentiation.

@@

\emph{\textbf{P53\textsuperscript{R172H} and P53\textsuperscript{R245W} cell lines are refractory to BRAF/MEK inhibition}}

Given that previous research has determined that p53 mutations can confer resistance to both chemotherapies and pathway-targeted therapies (Rusch \emph{et al.}, 1995; Huang \emph{et al.}, 2011), we next questioned whether our p53 alterations would be sufficient to drive therapeutic resistance in BRAF\textsuperscript{V600E}-driven cell lines exposed to pathway-targeted therapeutics commonly administered to BRAF\textsuperscript{V600E} mutant patients, such as the combination of dabrafenib (BRAFi) plus trametinib (MEKi) (Planchard \emph{et al.}, 2016). Cell lines were derived from \emph{Braf\textsuperscript{CAT/+}} murine tumors expressing either \emph{Trp53\textsuperscript{flfl}}, \emph{Trp53\textsuperscript{R172H/R172H}}, or \emph{Trp53\textsuperscript{R245W/R245W}}) 14 weeks after infection with ad5-Spc-Cre. Cell lines expressing either \emph{Trp53\textsuperscript{R172H/R172H}} or \emph{Trp53\textsuperscript{R245W/R245W}} were completely resistant to single agent dabrafenib and resistant to the combination of dabrafenib plus trametinib (Figure \ref{fig:13} A,B). Of note, \emph{Trp53\textsuperscript{R172H/R172H}} and \emph{Trp53\textsuperscript{R245W/R245W}} cell lines express increased ERK1/2 activity, as evidenced by increased phosphorylated ERK1/2, both with and without Dabrafenib and Trametinib (Figure \ref{fig:13} E). Moreover, \emph{Trp53\textsuperscript{R172H/R172H}} mice retain a larger tumor burden after a week 4 treatment of MAPK blockade \emph{in vivo} (Figure \ref{fig:13} E). To gain relevance to human lung cancer, we infected BRAF(V600E) mutant HCC364 cells with lentivirus expressing human cancer hotspot mutant p53\textsuperscript{R175H}. p53\textsuperscript{R175H} expressing cells were more proliferative and more refractory than parental HCC364 cells (Figure \ref{fig:13} F and G).Collectively, these \emph{in vitro} and \emph{in vivo} data suggest missense mutant p53\textsuperscript{R175H} and its murine homologue p53\textsuperscript{R172H}, as well as p53\textsuperscript{R245W} can support cancer cell viability in the face of BRAF+MEK blockade that cannot be achieved by p53-null mutations.

\emph{\textbf{P53\textsuperscript{R172H} and P53\textsuperscript{R245W} alters the transcriptome of BRAF(V600E) lung tumor cell lines}}

Due to our observations that mutant p53\textsuperscript{R172H} drives more aggressive cancer phenotypes such as increased growth, dedifferentiation, and resistance to FDA-approved therapeutics, we reasoned that this would be reflected in a change in the transcriptional landscape. This is due in part, by the transcriptional trans-activation potential by mutant p53. RNA sequencing of cell lines generated from p53-mutant mouse lung tumors do reveal transcriptomic differences between both p53\textsuperscript{R172H} and p53\textsuperscript{R245W} cells compared to p53-null cell lines (Figure \ref{fig:14} 4 A). Interestingly, genes involved in cell cycle regulation, epithelial-to-mesenchymal transition (EMT), and MYC pathway are upregulated in both p53\textsuperscript{R172H} and p53\textsuperscript{R2245W} cell lines (Figure 4 \ref{fig:14} B). These results suggest that mutant-p53 is affecting transcription of genes involved in pro-growth signaling pathways.

@@

\begin{figure}
\hypertarget{fig:12}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p53_2.png}
\caption{\textbf{Missense mutant-p53 alters lung adenocarcinoma differentiation status and accumulation of mutant-p53 correlates with p19 ARF expression.} \textbf{(A)} Panel representative IHC sections stained with SPC,NKX2-1, p53 and HMGA2. \textbf{(B)} H-score of individual tumors. \textbf{(C)} Association between mutant-p53 R172H accumulation and increased malignancy and dedifferentiation. \textbf{(D)} serial section of P53-HI and pERK1/2-HI tumors compared to adjacent low. \textbf{(E)} Colocalization of p19/ARF and p53 in p53-mutant tumors}\label{fig:12}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p53_1.png}
\caption{\textbf{This is the title of Figure} this is the text.}\label{fig:04}
}
\end{figure}

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p53_2.png}
\caption{\textbf{This is the title of Figure} this is the text.}\label{fig:04}
}
\end{figure}

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p53_3.png}
\caption{\textbf{This is the title of Figure} this is the text.}\label{fig:04}
}
\end{figure}

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p53_4.png}
\caption{\textbf{This is the title of Figure} this is the text.}\label{fig:04}
}
\end{figure}

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p53_5.png}
\caption{\textbf{This is the title of Figure} this is the text.}\label{fig:04}
}
\end{figure}

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p536.png}
\caption{\textbf{This is the title of Figure} this is the text.}\label{fig:04}
}
\end{figure}

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p537.png}
\caption{\textbf{This is the title of Figure} this is the text.}\label{fig:04}
}
\end{figure}

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p538.png}
\caption{\textbf{This is the title of Figure} this is the text.}\label{fig:04}
}
\end{figure}

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p539.png}
\caption{\textbf{This is the title of Figure} this is the text.}\label{fig:04}
}
\end{figure}

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p5310.png}
\caption{\textbf{This is the title of Figure} this is the text.}\label{fig:04}
}
\end{figure}

@@

@@

history of scRNA seq
cell types of lung tumorigenesis
tumor heterogenity
response to D+T

@@

Adenoviral vectors and lentiviral vectors introduced in the lungs of GEMMs offer temporal control and can be titrated to acheive an optimal level of tumor numerosity. However, only lentiviral vectors can be used to barcode tumors because they integrate into the genome while adenoviruses do not(DuPage \emph{et al.}, 2009; Rogers \emph{et al.}, 2017).

@@

\hypertarget{chapter-three-1}{%
\subsection{Chapter Three}\label{chapter-three-1}}

\hypertarget{conclusions}{%
\subsection{Conclusions}\label{conclusions}}

\hypertarget{future-prespectives}{%
\subsection{Future prespectives}\label{future-prespectives}}

single cell at later time points
metastasis
p53 homologous recombination
mechanisms of setd2 loss
complex gene-drug interactions
copy number variation
requirement for p19 for p53 mutant stability

>>>>

This manuscript
(\href{https://scherzerthesis.github.io/thesis/v/cacf67ddbb3ecfa8a81d823abdfbeafd930f1848/}{permalink})
was automatically generated
from \href{https://github.com/scherzerthesis/thesis/tree/cacf67ddbb3ecfa8a81d823abdfbeafd930f1848}{scherzerthesis/thesis@cacf67d}
on March 30, 2022.

@@

\textbf{P53-mutant Alleles Differentially Alter BRAF (V600E) Lung Tumorigenesis}

@@

\textbf{Mutant-p53 Accumulation is Associated with Dedifferentiation and p19 Induction}
Increased malignancy in lung adenocarcinoma is often associated with altered states of tumor cell identity and dedifferentiation (van Veen \emph{et al.}, 2019; Marjanovic \emph{et al.}, 2020). Since LUADs often arise from AT2 cells, we surveyed the expression of AT2 cell identifiers, such as NKX2-1, a master regulator of AT2 transcriptional programs, and SPC, a marker of differentiated AT2 cells (Marjanovic \emph{et al.}, 2020). Quantification of over 50 tumors from \emph{Braf\textsuperscript{V600E/+}} mice with either \emph{Trp53\^{} flfl\^{}}, \emph{Trp53\textsuperscript{R172H/R172H}}, and \emph{Trp53\textsuperscript{R245W/R245W}} revealed that \emph{Trp53\textsuperscript{R172H/R1722H}} mice had significantly more tumors that stained negative for NKX2-1 and SPC (Figure \ref{fig:12} A an B). This finding strongly indicates that mutant-p53 tumors are dedifferentiated. Since HMGA2 expression is associated with LUAD dedifferentiation and worse patient prognosis, we performed IHC for HMGA2 (Cite). We noted an increase in HMGA2-positive tumors in both \emph{Trp53\textsuperscript{R172H/R172H}}, and \emph{Trp53\textsuperscript{R245W/R245W}} tumors compared to \emph{Trp53\textsuperscript{fl/fl}} (Figure \ref{fig:12} A,B). Thus, this data taken together reveals that Trp53R172H/R172H mice exhibit both a higher tumor burden and worse overall survival than \emph{Trp53\textsuperscript{fl/fl}}, but the tumors developed in these mice are more likely to exhibit malignant progression and dedifferentiation.

@@

\textbf{P53\textsuperscript{R172H} and P53\textsuperscript{R245W} cell lines are refractory to BRAF/MEK inhibition}
Given that previous research has determined that p53 mutations can confer resistance to both chemotherapies and pathway-targeted therapies (Rusch \emph{et al.}, 1995; Huang \emph{et al.}, 2011), we next questioned whether our p53 alterations would be sufficient to drive therapeutic resistance in BRAF\textsuperscript{V600E}-driven cell lines exposed to pathway-targeted therapeutics commonly administered to BRAF\textsuperscript{V600E} mutant patients, such as the combination of dabrafenib (BRAFi) plus trametinib (MEKi) (Planchard \emph{et al.}, 2016). Cell lines were derived from \emph{Braf\textsuperscript{CAT/+}} murine tumors expressing either \emph{Trp53\textsuperscript{flfl}}, \emph{Trp53\textsuperscript{R172H/R172H}}, or \emph{Trp53\textsuperscript{R245W/R245W}}) 14 weeks after infection with ad5-Spc-Cre. Cell lines expressing either \emph{Trp53\textsuperscript{R172H/R172H}} or \emph{Trp53\textsuperscript{R245W/R245W}} were completely resistant to single agent dabrafenib and resistant to the combination of dabrafenib plus trametinib (Figure \ref{fig:13} A,B). Of note, \emph{Trp53\textsuperscript{R172H/R172H}} and \emph{Trp53\textsuperscript{R245W/R245W}} cell lines express increased ERK1/2 activity, as evidenced by increased phosphorylated ERK1/2, both with and without Dabrafenib and Trametinib (Figure \ref{fig:13} E). Moreover, \emph{Trp53\textsuperscript{R172H/R172H}} mice retain a larger tumor burden after a week 4 treatment of MAPK blockade \emph{in vivo} (Figure \ref{fig:13} E). To gain relevance to human lung cancer, we infected BRAF(V600E) mutant HCC364 cells with lentivirus expressing human cancer hotspot mutant p53\textsuperscript{R175H}. p53\textsuperscript{R175H} expressing cells were more proliferative and more refractory than parental HCC364 cells (Figure \ref{fig:13} F and G).Collectively, these \emph{in vitro} and \emph{in vivo} data suggest missense mutant p53\textsuperscript{R175H} and its murine homologue p53\textsuperscript{R172H}, as well as p53\textsuperscript{R245W} can support cancer cell viability in the face of BRAF+MEK blockade that cannot be achieved by p53-null mutations.

\textbf{P53\textsuperscript{R172H} and P53\textsuperscript{R245W} alters the transcriptome of BRAF(V600E) lung tumor cell lines}
Due to our observations that mutant p53\textsuperscript{R172H} drives more aggressive cancer phenotypes such as increased growth, dedifferentiation, and resistance to FDA-approved therapeutics, we reasoned that this would be reflected in a change in the transcriptional landscape. This is due in part, by the transcriptional trans-activation potential by mutant p53. RNA sequencing of cell lines generated from p53-mutant mouse lung tumors do reveal transcriptomic differences between both p53\textsuperscript{R172H} and p53\textsuperscript{R245W} cells compared to p53-null cell lines (Figure \ref{fig:14} 4 A). Interestingly, genes involved in cell cycle regulation, epithelial-to-mesenchymal transition (EMT), and MYC pathway are upregulated in both p53\textsuperscript{R172H} and p53\textsuperscript{R2245W} cell lines (Figure 4 \ref{fig:14} B). These results suggest that mutant-p53 is affecting transcription of genes involved in pro-growth signaling pathways.

\textbf{addiction to mutant-p53 expression}
\textbf{PDGFRA RNA data}
\textbf{Response to PDGFR stimuliation and inhibition}

@@

\begin{figure}
\hypertarget{fig:12}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p53_2.png}
\caption{\textbf{Missense mutant-p53 alters lung adenocarcinoma differentiation status and accumulation of mutant-p53 correlates with p19 ARF expression.} \textbf{(A)} Panel representative IHC sections stained with SPC,NKX2-1, p53 and HMGA2. \textbf{(B)} H-score of individual tumors. \textbf{(C)} Association between mutant-p53 R172H accumulation and increased malignancy and dedifferentiation. \textbf{(D)} serial section of P53-HI and pERK1/2-HI tumors compared to adjacent low. \textbf{(E)} Colocalization of p19/ARF and p53 in p53-mutant. tumors}\label{fig:12}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p53_5.png}
\caption{\textbf{This is the title of Figure} this is the text.**}\label{fig:04}
}
\end{figure}

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p536.png}
\caption{\textbf{This is the title of Figure} this is the text.**}\label{fig:04}
}
\end{figure}

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p537.png}
\caption{\textbf{This is the title of Figure} this is the text.**}\label{fig:04}
}
\end{figure}

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p538.png}
\caption{\textbf{This is the title of Figure} this is the text.**}\label{fig:04}
}
\end{figure}

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p539.png}
\caption{\textbf{This is the title of Figure} this is the text.**}\label{fig:04}
}
\end{figure}

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p5310.png}
\caption{\textbf{This is the title of Figure} this is the text.**}\label{fig:04}
}
\end{figure}

\begin{figure}
\hypertarget{fig:04}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/p5311.png}
\caption{\textbf{This is the title of Figure} this is the text.**}\label{fig:04}
}
\end{figure}

@@

\#\#Abstract

@@

@@

\textbf{table of genotypes/treatment/and cells collected}
Pre-sort
pre-batch correction
batch correction
cell types
tumor cells only
Feature plots of AT2 markers
feature plots of AT1 markers
Feature plots of P53 target genes
heatmap of P53 target genes
heatmap/violin plot of MAPK score

@@
